Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-09 src/article.tex
Move text directory structure up one directory
Revise results section

The Scaffolder software allows manual or computational joining of multiple
discontiguous nucleotide sequences together into a contiguous super sequence.
Simple plain-text configuration files are used to select the nucleotide
sequences in the scaffold and how they should be joined together. This is
called the ``scaffold file''. \pb

The scaffold file is read by the scaffolder software to determine how the
scaffold super sequence is generated. In addition to specifying which
nucleotides should be used to generate the draft super sequence, the scaffold
file allows the sequences to trimmed down to smaller sub-sequences and to
reverse complement the sequence if necessary. \pb

There may be unknown regions in the sequences used in the scaffold. Region of
nucleotide sequence the assembler was unable to determine the nucleotide
sequence. The process of genome finishing may involve determining the correct
nucleotide sequence for these regions then filling the gaps with the additional
correct sequence. The scaffold file specifically allows these inserts to be
added to scaffold. These inserts can be treated the same way as a larger
contig: trimmed and/or reversed to match the required gap region. \pb

When scaffolding a genome there may be regions of sequence approximate size
region may be estimated from mate-pair insert size or from mapping the contigs
to reference genome. These unknown regions are useful to join non-contiguous
sequences together by the estimated size of the region whilst still
highlighting areas in the genome to be resolved. These unknown regions can be
specified in the scaffold file for the above reasons. \pb

The original raw nucleotide sequence for each assembled contig or insert
sequence is maintained in a fasta file. The fasta file of contigs is a common
output for assemblers and as such should be readily available after assembly
without additional processing. The choice of the nucleotide sequence to use in
the scaffold is specified using the first space-delimited word from the fasta
header of the sequence. By maintaining the nucleotide sequences in a separate
file this preserves the original raw sequence in the fasta file. Furthermore
this separates the actual nucleotide sequence from the determination of how it
is selected and edited in the final genome sequence. \pb

@@

The scaffold file is written using `YAML', a data format designed to be both
machine readable and manually editable. An example of a scaffold file in YAML
format is shown in Figure 1. This example scaffold file illustrates the
available features in scaffolder. The basic layout of the scaffold file is as
a list - a series of entries where each entry defines a region of sequence to
be generated for the final super sequence. In the final super sequence each of
the entries are joined together to generate a single continuous sequences. \pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. Line 2 begins
with a dash character `-' denoting the first entry in the sequence scaffold.
Line 3 is indented by two spaces. Whitespace is used to delimit attributes in
the file. Lines with an incremented level of white space are considered linked
to the first preceding non-indented line. \pb

The ``sequence'' tag used on line 3 indicates that this region is in the
scaffold will contain nucleotide sequence from the fasta file. The following
line indented by two spaces and indicating an attribute of the sequence entry
defined on line 3. In this case the ``source'' keyword identifies the name of
the sequence to be inserted into the scaffold. As described above the first
space-delimited word of the fasta header should be used to identify the
sequence. Taken together lines 2-4 describe the first entry in the scaffold is
a simple sequence which the identifier `sequence'. \pb

@@

Line 5 has no whitespace indentation and begins with a dash and therefore
describes the next entry in the scaffold. This entry is identified by the
``unresolved'' tag, and specifies a region of unknown nucleotides but known
length. The following line specifies the size of this unknown region. Lines 5-6
therefore insert a region of 20 `N' characters in the super sequence directly
following the region defined by the first entry on lines 2-5. \pb

@@

The last entry adds the sequence `sequence2' in the scaffold file. This entry
demonstrates three further attributes for a sequence entry.The `start' and
`stop' tags are used to trim the sequence used in the scaffold to these
coordinates inclusively. The ``reverse'' tag is used to reverse complement this
region in the final super sequence. \pb

@@

Lines 16-19 illustrate an insert using similar attributes to that of a sequence
entry. The reverse, start and stop tags trim the inserted sequence and the
`source' tag identifies the corresponding fasta sequence. The ``open'' and
``close'' tags on lines 20-21 determine where the insert is added to the
enclosing sequence.  These coordinates inclusively replace the region within
these coordinates with the specified fasta sequence. The open and close
positions need not be the same size as the inserted sequence. \pb

The next insert on lines 23-24 specifies only the open position. This
illustrate that only the one of the `open' or `close' tags is required.  When
only open or close tag is used then the opposing position is calculated from
the length on the insert fasta sequence. This allows inserts to partially fill
gap regions in sequences. \pb

@@

The sequence entry in the scaffold file allows multiple sequence-editing
attributes which when applied in different combinations will result in
different output sequence. An example is reverse complementing the sequence
then following adding sequence inserts. Performing these actions in the reverse
order will produce a difference output sequence. This section therefore
outlines the scaffolder protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts these are reverse sorted by close position. The insert with the
largest close position is first in the list and inserts with the smallest
close position is last. This preserves the original indented coordinates for
insert during the process of adding inserts to the sequence. \pb

\paragraph{Update sequence with insert:} Each insert sequence is processed
according to any specified attributes. The insert is added to the enclosing
sequence in the region specified by the open and close coordinates. \pb

If the insert added to the sequence is a different length than the region
replaced the sequence start and stop co-ordinates are updated. For example if
a 7bp insert replaced a 5bp region the stop coordinate of the enclosing
sequence would be incremented by the 2bp difference. \pb

\paragraph{Trimming sequence:} If the sequence is has start or stop sequences
specified these are used to trip the sequence to these co-ordinates
inclusively. \pb

\paragraph{Reverse complement sequence:} If the ``reverse'' keyword is
specified the sequence is reverse complemented. This is final step in
processing the sequence with results in the sequence being returned. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (the
application programming interface or API) allowing the scaffold to be
manipulated programmatically. This allows scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as using Rake.
\pb

Scaffolder also provides a command line interface (CLI) through a scaffolder
binary. This the scaffold sequence to be produced and validated without
requiring any Ruby knowledge. The scaffolder CLI behaves as a standard Unix
tool - returning appropriate exit statuses and providing appropriate manual
pages. The use of both these scaffolder interfaces is outlined in the following
sections. \pb

@@

Scaffolder relies on both the ruby programming language and the RubyGems
package management system to work. These are either already present easily
installed on recent distributions of Mac OSX or Linux. The presences of these
tools can be tested by typing the following on the command line. \pb

@@

If both these commands return without error this indicates the necessary
system requirements to install and run scaffolder. The scaffolder API and CLI
can then be installed through RubyGems with the following command.

@@

To ensure that the scaffolder CLI is accessible the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed in
~/.gem this can be done by adding the following to the ~/.bashrc file. This
should be adjusted for other shells such as Zsh etc..

@@

The Scaffolder CLI provides functions to validate the scaffold file and build the draft scaffold sequence. Assuming scaffolder has been installed and the
binary is available in the command line path the following command should run without error and list the available functions. \pb

@@

Help on a specific command can be found by typing `help' followed by the name
of the command. This will return the corresponding Unix manual page. The
following example will return the manual page for the sequence command. \pb

@@

The ``sequence'' command builds the fasta sequence from the scaffold and fasta
file. Running this command requires the locations of both the scaffold file and
the fasta file of sequences. This will then return the complete scaffold
sequence. \pb

@@

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the scaffolder website. The following code
snippet outlines how the API may be used to create a script to replicate the
function of the `scaffolder sequence' command described above. \pb

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable object as the next line illustrates the
trivial code required to generate the scaffold sequence. Further details of
the classes and method related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class. \pb

@@

The main aims of scaffolder are two fold. The first is produce easy to
install and simplifies the task of genome finishing. The second aim is to
allow to create a tool to facilitate reproducibility in the task of joining
large nucleotide sequences together. \pb

\subsubsection*{Simple to use}

Scaffolder is designed to be simple to use as possible. Assuming the existence
of Ruby and RubyGems, scaffolder can be installed in a single command line
step. Scaffolder uses a minimal and compact syntax to described the genome
scaffold. This syntax is simple to construct and edit whilst being succinct
and readable. This aims to allow creating a genome scaffold to relatively
simple to do in a common text editor. Furthermore scaffolder provides
extensive documentation both for the command line tools and for the Scaffolder
Ruby API. Therefore should any part of the scaffolder tool be unclear the
documentation should make an solution easy to find. \pb

\subsubsection*{Reproducible genome construction}

Constructing a genome sequence using a plain text scaffold file means the same
sequence is always output for the same file. Joining large nucleotide
sequences manually cannot however be reliably reproduced. The layout of the
scaffold file furthermore provides a human readable record of how the sequence
was constructed. \pb

Configuring the final sequence in the scaffold file means the build is easier
to edit once already constructed compared with working on the sequence alone.
The YAML text format is also much easier to compare differences using standard
text comparison tools such as diff. This makes scaffold files easier to store
in version control systems. This affords genome scaffolding the same benefits
enjoyed by software developers. \pb

@@

\section*{Acknowledgements} %{{{1
  \ifthenelse{\boolean{publ}}{\small}{}
  Text for this section \ldots

>>>>

The Scaffolder software outlined here is designed to allows manual or
computational joining of nucleotide sequences together into a single contiguous
super sequence. Plain-text files written using the YAML syntax files are used
to specify the nucleotide sequences in the scaffold and how they should be
joined together. This file is described as the ``scaffold file''. \pb

The scaffold file is read by the scaffolder software to determine how the
scaffold sequence should be generated. In addition to specifying which
nucleotides should be used to generate this draft sequence, the scaffold file
allows the sequences to trimmed down to smaller sub-sequences and reverse
complemented if necessary. \pb

The process of genome finishing may involve producing additional nucleotide
sequence to fill gaps in the scaffold. The scaffold file format allows these
additional insert sequences to be added to scaffold to bridge the gaps inside
contigs. These inserts can also be treated in a similar way to larger contig
sequences: trimmed and/or reverse complemented to match the corresponding gap
region. \pb

In the process of build a genome scaffold distances between contigs may be
estimated from paired read insert size or from mapping the contigs to reference
genome. These unknown gap regions are useful to join non-contiguous sequences
together by the estimated size of the region whilst still highlighting areas in
the scaffold to be resolved. The scaffold file also allows for the
specification of these unknown regions. \pb

The scaffold file specifies how the contigs nucleotide sequences are joined
together, however the sequences are not present in the scaffold file. These
nucleotide sequences are instead maintained in a fasta file. A fasta file of
contigs is a common output for assemblers and as such should be readily
available without further processing. Nucleotide sequence are specified in the
scaffold using the first space-delimited word from the fasta header.
Maintaining the nucleotide sequences in a separate file preserves the original
unedited sequence and furthermore separates raw sequence from the markup
describing how it should be used in the scaffold. \pb

@@

The scaffold file is written using the YAML syntax. YAML is a data format
designed to be both machine readable and easy to edit manually. An example
scaffold file is shown in Figure 1. This file illustrates the text attributes
used to describe a genome scaffold. The basic layout of the scaffold file is as
a list - a series of entries where each entry insert to a region of sequence in
the scaffold super sequence. \pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. Line 2 begins
with a dash character `-', dashes are used to denote entries in a YAML list.
This first line therefore indicates indicate the first entry in the scaffold.
Line 3 is indented by two spaces where whitespace is used to group similar
attributes together. Lines indented by whitespace are linked to the first
non-indented preceding line. \pb

The ``sequence'' tag on line 3 indicates this scaffold entry corresponds to an
entry in the fasta file. The following line, indented by two spaces, indicates
the name of the fasta sequence using the ``source'' keyword. As described above
the first space-delimited word of the fasta header should be used to identify
which sequence is used. Therefore lines 2-4 describe the first entry in the
scaffold as a simple sequence using the fasta sequence named `sequence'. \pb

@@

Line 5 has no whitespace indentation and begins with a dash therefore
describing the next entry in the scaffold. This entry is identified by the
``unresolved'' tag, and specifies a region of unknown nucleotides but known
length. The following line specifies the size of this unknown region. Lines 5-6
therefore insert a region of 20 `N' characters in the scaffold. \pb

@@

The last entry adds the fasta entry `sequence2' to the scaffold. This entry
demonstrates how sequence may be edited when added to the scaffold. The `start'
and `stop' tags trim the sequence to these coordinates inclusively. The
``reverse'' tag reverse complements this region. \pb

@@

Lines 16-19 illustrate an insert using similar attributes to that of a sequence
entry. The reverse, start and stop tags trim the inserted sequence and the
`source' tag identifies the corresponding fasta sequence. The ``open'' and
``close'' tags on lines 20-21 determine where the insert is added to the
enclosing sequence.  These coordinates inclusively replace the sequence region
within these coordinates with the specified insert fasta sequence. The open and
close positions need not be the same size as the replaced region. \pb

The next insert on lines 23-24 is specified by only the open position tag. This
illustrates that only the one of the `open' or `close' tags is required.  When
only one of these tag is used the corresponding opposing coordinate position is
calculated from the length in the insert fasta sequence. This allows inserts to
bridge into and partially fill gap regions. \pb

@@

The sequence entry in the scaffold file allows multiple sequence-editing
operations to be applied. If applied in different combinations, these
operations could result in different output sequence. An example would be
reverse complementing the sequence then trimming the start of the sequence by
five nucleotides. Performing these actions in the reverse order would produce
a difference output sequence. This section therefore outlines the scaffolder
protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts these are reverse sorted by the insert close coordinate. The insert
with the largest close position is first in the list and the insert with the
smallest close position is last. This preserves the original indented
coordinates for each insert during the process of adding inserts to the
sequence. \pb

\paragraph{Update sequence with insert:} Each insert sequence is processed
according to any specified attributes. The insert is then added to the
enclosing sequence in the region specified by the open and close coordinates.
\pb

If the insert added to the sequence is a different size compared to the region
it replaces then the sequence start and stop co-ordinates are updated.  For
example if a 7bp insert replaced a 5bp region in the sequence the stop
coordinate of the enclosing sequence would be incremented by the 2bp
difference. As inserts are added by reverse close position order the remaining
inserts are unaffected by this change in the sequence size. \pb

\paragraph{Trimming sequence:} If the sequence has start or stop sequences
specified these are used to trim the sequence to the specified length
inclusively. \pb

\paragraph{Reverse complement sequence:} If the reverse tag is specified the
sequence is reverse complemented. This is final step in processing the sequence
and results in the sequence being returned for inclusion in the scaffold. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API)
allowing the scaffold to be manipulated programmatically. This allows
scaffolder to be integrated into existing genomics workflows or used with Ruby
build tools such as Rake. \pb

In addition Scaffolder provides a command line interface (CLI) through the
scaffolder binary. This allows the scaffold sequence to be produced and
validated without requiring any Ruby experience. The scaffolder CLI behaves as
a standard Unix tool - returning appropriate exit statuses and providing
appropriate manual pages. The use of both these scaffolder interfaces is
outlined in the following sections. \pb

@@

Scaffolder relies on both the Ruby programming language and the RubyGems
package management system. These are either already present or easily
installed on recent distributions of Mac OSX or Linux. The presence of these
tools can be tested by typing the following on the command line. \pb

@@

If both these commands return without error this indicates the necessary
system requirements to install and run scaffolder. The scaffolder API and CLI
can then be installed through RubyGems with the following command. \pb

@@

To ensure that the scaffolder CLI is accessible, the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed in
~/.gem this can be done by adding the following to the ~/.bashrc file. The
corresponding login shell should be updated for other shells such as Zsh, etc..
\pb

@@

The Scaffolder CLI provides functions to validate the scaffold file and build
the draft scaffold sequence. Assuming scaffolder has been installed and the
binary is available in the command line path the following command should run
without error and list the available scaffolder functions. \pb

@@

Help on a specific command can be found by typing `help' followed by the name
of the command. This returns the corresponding Unix manual page. The following
example will return the manual page for the sequence command. \pb

@@

The ``sequence'' command builds the fasta sequence from the scaffold and fasta
file. Running this command requires the locations of both the scaffold file and
the fasta file to be passed as argument. This will then return the complete
scaffold sequence in fasta format. \pb

@@

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the scaffolder website. The following code
snippet outlines how the API may be used to create a script to replicate the
function of the `scaffolder sequence' command described above in the CLI
section. \pb

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable object and the next line illustrates the
trivial code required to generate the scaffold sequence. Further details of the
classes and method related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class. \pb

@@

The aim of scaffolder are twofold. The first aim is to provide scaffolding
software which is easy to install and simplifies the task of genome finishing.
The second aim is to facilitate reproducibility when joining large nucleotide
sequences together to create a genome scaffold. \pb

\subsubsection*{Simple to use} %{{{3

Scaffolder is designed to be simple to use as possible. Assuming the existence
of Ruby and RubyGems, scaffolder can be installed in a single command line
step. This should negate the possible barriers to use in requiring manual
compilation of the source code. \pb

Scaffolder uses a minimal and compact syntax to described the genome scaffold.
This syntax is simple to construct and edit whilst being succinct and readable.
The aim of this is genome scaffold creation to relatively simple in a common
text editor. Furthermore scaffolder provides extensive documentation both for
the command line tools and for the Scaffolder Ruby API. Therefore should any
part of the scaffolding process be unclear the documentation should make an
solution easy to find. \pb

\subsubsection*{Reproducible genome construction} %{{{3

Constructing a genome sequence using a text file to specify the scaffold means
the same sequence is always output for the same file. Joining large nucleotide
sequences manually cannot however be reliably reproduced. The layout of the
scaffold file furthermore provides a human readable record of how the sequence
was constructed. The same cannot be said for large block of nucleotide
sequences comprised of many source contigs. \pb

Configuring the final sequence in the scaffold file means the build is easier
to edit once already constructed compared with working with an already drafted
sequence. The YAML text format is also facilitates easier comparison of
differences between scaffold build using standard text comparison tools such as
diff. This makes scaffold files easier to store in version control systems and
affords genome finishers similar benefits enjoyed by software developers. \pb

@@

\section*{Acknowledgements} %{{{1
  \ifthenelse{\boolean{publ}}{\small}{}

This work was supported by the National Institute for Health: IDeA Network of
Biomedical Research Excellence (KY-INBRE) grant: NIH 2P20 RR016481-09 and the
NIH R15 AREA Program grant: NIH R15GM079775.
